\thetitle

### 2021-02-21 writing/challenges.tex
describe challenges systematically, factor out the implementation challenge

The challenge of evaluating blame is a methodological one. Unlike most research
on programming languages, the question seems to call for empirical research
similar to that of human-computer interaction. At the same time, a significant
evaluation result demands a large number of cases yielding a serious amount of
data. As~\citet{lksfd-popl-2020} recently demonstrated, the way around this
dilemma is to postulate the equivalent of a {\it homo economicus\/} for
programming languages---the {\em rational programmer\/}---and to implement this
rational programmer as an algorithm.

In the context of gradual typing, the rational programmer translates the
Wadler-Findler slogan into a debugging method, searching for the source of the
impedance mismatch in an incremental fashion. Measuring this simulated
programmer's behavior on a large number of debugging sessions yields data that
is similar to empirical data collected in a human-focused discipline.

A derivative challenge is the problem of simulating the impedance mismatches
between typed and untyped pieces of code that programmers may accidentally create
and that the run-time checks of academic systems catch. In other words, the
experiment calls for a collection of type mistakes in mix-typed programs that is
representative of those ``in the wild.'' Unfortunately no such collection exists,
and with good reason; the kind of mistakes the rational programmer targets are
easily detected by unit tests and, even if their sources take some time to find,
these mistakes do not make it into code repositories with appropriate commit
messages.  Hence the experimenter must generate a sufficiently large and
representative corpus of type mistakes for a suitable collection of benchmark
programs.  Following~\citet{lksfd-popl-2020} again, the solution is to synthesize
mistakes using mutation analysis~\cite{lipton1971fault, demillo1978hints,
jia2011analysis}.  The mutation analysis of Lazarek et al, is, however, useless
for creating impedance mismatches.  While mutation analysis traditionally aims to
inject bugs that challenge test suites, it discards those that yield ill-typed
mutants as \emph{incompetent}.  Indeed, mutation analysis frameworks are
fine-tuned to avoid them---and yet it is precisely those that are needed for
evaluating blame assignment strategies.

Thus, an experimental analysis of blame needs a mostly new set of mutators.
Roughly speaking, the new mutators inject type errors into fully typed programs.
Applying such a mutator to any typed component produces a mutatnt component.  A
debugging scenario results from removing the types from the mutatnt. For the
design of such mutators, the authors relied on their own extensive programming
experience though not without discovering a major pitfall: some of their original
mutators systematically produced programs that immediately revealed the source of
the impedance mismatch. About two thirds of the original mutators survived this
first scrutiny testing. All of the remaining ones yield interesting debugging
sessions.

A somewhat coincidental challenge is the disparity of gradually typed
languages. While the approaches are well understood in theory, they come with
vastly different type systems, accommodate seemingly incompatible language
features, and apply to radically different coding styles in the untyped world.
To eliminate the effect of these differences on its results, the application of
an evaluation method to different gradual typing systems must use the same test
programs and the same type system. Otherwise, results for different gradual
typing systems do not share a common baseline and, therefore, do not lend support
to any conclusions about whether the value of blame changes from one system to
another. For instance, the case study presented here builds on Greenman and
Felleisen's Transient variant for functional Typed Racket~\cite{gf-icfp-2018} to
obtain a Transient semantics that is comparable with Typed Racket's default
Natural one. This Transient variant accommodates varied linguistic features and
library calls in a way that is faithful to the Transient
semantics~\cite{vss-popl-2017}.

Beyond the sheer implementation issue, the development of a method must also
address the problem that the three semantic variants---Natural, Transient, and
Erasure---come with three different blame assignment strategies.  While Natural
assigns blame to {\em one\/} component, Transient assigns blame to a sequence of
components. The Erasure semantics does not blame a component per se, but it comes
with an exception location and a stack trace.

Supporting these strategies is {\em not\/} just an issue of language
implementation but affects greatly the behavior of the rational
programmer.  Specifically, the rational programmer must come with
\emph{modes} that represent the alternative interpretations of blame.  For
instance, a mode of the rational programmer using Transient may pick the
oldest element of a blame set as the next component to add types because
it corresponds to the earliest point in the program's evaluation that can
discover an impedance mismatch.  Another mode may opt to treat the blame
set as another form of a stack trace and add types to its newest element.
When both modes are equally successful in locating an impedance mismatch,
the method needs to take into account the rational programmer's effort
to answer which mode is the most effective.

%Another mode does not distinguish between the elements of the
%blame set and has the rational programmer type all blamed components
%at once. Because this second kind of step 
%is more laborious than the first one, the analysis of the results must take 
%into account the rational programmer's effort.

Besides the blame information they produce, the three gradual typing
systems also differ in the runtime type checks they perform. Therefore an
application of the evaluation method to a gradual typing system has to
factor in the effect of their different type checks to determine whether blame helps the
rational programmer find the cause of an impedance mismatch. A solution to this issue
is yet another mode for the rational programmer that uses information only
from the underlying languages exceptions to select the next component(s) to type.
This \emph{exception} mode simulates a rational programmer that ignores
blame and serves as the baseline for blame's value.

The final big challenge is the explosion of debugging scenarios that results from
the combination of mutation-based scenario generation and mode-based analysis.
Specifically, the experiment demands a large amount of computational resources
due to the multiplicative effect of mixing the different gradual type systems,
the number of debugging modes for the rational programmer, and the large number
of mutants.  A practical application of the method has no option but to sample
the space of scenarios, carefully ensuring that it still produces reproducible
conclusions.

\smallskip

Overcoming these challenges needs a systematic approach:

%% for doing so correctly
%% MF: what is correct about this recipe?

\begin{itemize}

\item a single-language implementation of all three gradual type systems (section~\ref{sec:landscape});

\item a suitable set of mutators for the generation of a large set of debugging
  scenarios with interesting type-level errors (section~\ref{sec:mutate});

\item an algorithmic simulation of the rational programmer on these debugging
  scenarios with sufficient modes to make meaningful comparisons between modes of
  the same or different systems (section~\ref{sec:rational});

\item a careful sampling of the experimental space to obtain a computationally
  feasible and statistically sound experiment (section~\ref{sec:sample}).

\end{itemize}

>>>>

The general challenge of evaluating blame is a methodological one. Unlike most
current research on programming languages, the question seems to call for
empirical research similar to that of human-computer interaction. At the same
time, a significant evaluation result demands a large number of cases yielding a
serious amount of data. As~\citet{lksfd-popl-2020} recently demonstrated, the
way around this dilemma is to postulate the equivalent of a {\it homo
economicus\/} for programming languages---the {\em rational programmer\/}---and
to implement this rational programmer as an algorithm.

In the context of gradual typing, the rational programmer translates the
Wadler-Findler slogan into a debugging method, searching for the source of the
impedance mismatch in an incremental fashion. Measuring this simulated
programmer's behavior on a large number of debugging sessions yields data that
is similar to data collected in an empirical, human-facing manner.

The {\em first derivative challenge\/} is to find a representative collection of
impedance mismatches between typed and untyped pieces of code that programmers
may accidentally create and that the run-time checks of academic systems
catch. In other words, the experiment calls for a collection of type mistakes in
mix-typed programs that is representative of those ``in the wild.''
Unfortunately no such collection exists, and with good reason. The kind of
mistakes needed are typically detected by unit or integration tests; even if it
takes some time to find their sources, these mistakes do not make it into code
repositories with appropriate commit messages.

Following~\citet{lksfd-popl-2020} again, the alternative is to {\em generate a
sufficiently large and representative corpus of type mistakes \/} using mutation
analysis~\cite{lipton1971fault, demillo1978hints, jia2011analysis}.  The
mutation analysis of Lazarek et al, is, however, useless for creating impedance
mismatches.  While mutation analysis traditionally aims to inject bugs that
challenge test suites, it discards those that yield ill-typed mutants as
\emph{incompetent}.  Indeed, mutation analysis frameworks are fine-tuned to
avoid them---and yet it is precisely those that are needed for evaluating blame
assignment strategies.

Thus, an experimental analysis of blame needs a mostly new set of mutators.
Roughly speaking, the new mutators inject type errors into fully typed programs.
Applying such a mutator to any typed component produces a mutatnt component.  A
debugging scenario results from removing the types from the mutatnt. For the
design of such mutators, the authors relied on their own extensive programming
experience though not without discovering a major pitfall: some of their
original mutators systematically produced programs that immediately revealed the
source of the impedance mismatch. About two thirds of the original mutators
survived this first scrutiny testing. All of the remaining ones yield
interesting (see section~{sub:interesting}) debugging sessions.

The {\em second derivative challenge\/} concerns the differences among the blame
assignment of the three semantic variants.  While Natural assigns blame to {\em
one\/} component, Transient assigns blame to a sequence of components. The
Erasure semantics does not blame a component {\it per se\/}, but it comes with
an exception location and a stack trace, which implicitly suggests parties to
blame.  The different strategies greatly affect the behavior of the rational
programmer (and real programmers, too).

One way to overcome this challenge is to equip the rational programmer with
\emph{modes} that represent the alternative interpretations of blame.  For
instance, a mode of the rational programmer using Transient may pick the oldest
element of a blame set as the next component to add types because it corresponds
to the earliest point in the program's evaluation that can discover an impedance
mismatch.  Another mode may opt to treat the blame set as another form of a
stack trace and add types to its newest element.  When both modes are equally
successful in locating an impedance mismatch, the method can measure the
rational programmer's effort to answer which mode is the most effective.

Different blame information is also due to differences in the run-time checks
that the three gradual typing systems perform. This observation suggests another
mode for the rational programmer, namely, a mode that ignores the blame
information completely and instead inspects only the information in the
exceptions of the safety checks in the underlying languages.  This
\emph{exception} mode serves as the baseline for blame's value; if the
programmer in this mode ``beats'' the others, blame systems might be useless.

The {\em third challenge\/} is the explosion of debugging scenarios that results
from the combination of mutation-based scenario generation and mode-based
analysis. All three factors---three different gradual typing systems, the large
number of mutants, and the number of debugging modes---contribute possibly
useful debugging data in multiplicative manner. Hence, carried out naively, the
experiment would demand an unacceptably large amount of computational resources.
A practical execution has no option but to sample the space of scenarios,
carefully ensuring that it still produces reproducible conclusions.

Overcoming these challenges needs a systematic approach: 
\begin{itemize}

\item a suitable set of mutators for the generation of interesting type-level
 errors (section~\ref{sec:mutate});

\item an algorithmic simulation of the many modes of the rational programmer on
 the corresponding debugging scenarios (section~\ref{sec:rational});

\item a careful sampling of the space (section~\ref{sec:sample}).
\end{itemize}
Getting all three aspects right is necessary to obtain a computationally
feasible and statistically sound experiment

{\bf Note} A somewhat coincidental challenge is the disparity of gradually typed
languages. While the three approaches to blame are well understood in theory,
they have been implemented for rather different untyped languages and with
vastly different type systems. To eliminate the effect of these differences, it
is necessary to implement all three approaches in a single language so that an
application of an evaluation method can use the same test programs, the same
type system, and hence the same debugging scenarios.

The results in this paper are for a system that builds on Greenman and
Felleisen's effort comparing the performance of Natural and Transient in
Racket~\cite{gf-icfp-2018}. The extended system uses the type system of Typed
Racket and comes with a Transient semantics that is comparable with the Natural
one. The Transient semantics of the system also realizes the design of the blame
assignment system proposed by~\cite{vss-popl-2017}. The Erasure semantics drops
the type information and runs the programs as if they were plain Racket
programs. The next section explains this system in detail.
